---
title: "Basic Exploration"
author: "Simon Tavener and Stu Field"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output:
  html_notebook:
    toc: true
    toc_float:
      collapsed: false
    code_folding: show
    number_sections: true
ratio: '9:16'
tables: yes
fontsize: 12pt
---


# Introduction

This is a very simple streamlined set of exmple code to describe some basic summary statistics and graphs to facilitate preliminary data exploration.


# Data

We will be using a few data sets during the course, one of the classic statistical learning data sets is the *iris* data set which comes with any standard installation of R. This famous (Fisher's or Anderson's) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are *Iris setosa*, *versicolor*, and *virginica*.
```{r}
?iris        # open help file for the object
dim(iris)        # get dimensions of data
class(iris)  # a data frame class object
```

The first goal for any analysis is to understand the dimensions and structure of the data set you are working with. Here are some simple commands to understand the general structure of your data set, typically in the form of a **data.frame** object:
```{r}
dim(iris)            # dimensions (rows, columns)
str(iris)            # structure of the data frame object
iris[1:5,]           # echo the first 5 rows
names(iris)          # what are the column names
sapply(iris, class)  # what are the classes of each column (very important!); feature vs. class data
table(iris$Species)  # once you know which column contains the classes, what are the class counts?
```

The *summary* function is a special class of generic functions (called S3) that perform specific operations depending on the class object they are performing on ... in this case a "data.frame" object. The default is to "summarize" the columns, reporting a different summary for each:
```{r}
summary(iris)
```

However, it is often more useful from a classification perspective to know the feature means *by class*, which can be accomplished via *tapply*:
```{r}
sapply(iris[, -5], function(x)
   tapply(x, iris$Species, mean))   # feature means split by Species
```

There is a new syntax that has become quite popular in recent years, referred to simply as `tidy` data analysis, spearheaded by Hadley Wickham. Here is how you would get to the same answer using this syntax:

```{r}
library(tibble)
iris %>% as.tibble %>%    # make 'iris' a `tibble` data frame
  group_by(Species) %>%
  summarise_all(funs(mean))
```



# Splitting Data
All of the following commands produce exactly the same output:
```{r}
split(iris$Sepal.Length, iris$Species)
split(iris[["Sepal.Length"]], iris$Species)
split(iris[[1L]], iris$Species)
split(iris[[which(names(iris)=="Sepal.Length")]], iris$Species)
```

## Exercise: what does the following command return?
```{r, eval = FALSE}
split(iris, iris$Species)
```


----------------------

# Plotting

```{r}
par(mfrow=c(2, 2))                       # make 2x2 grid for plots
par(mgp=c(2, 0.75, 0), mar=c(3, 4, 3, 1))    # graphics settings; squeeze margins
lapply(names(iris)[-5], function(class)
   boxplot(split(iris[[class]], iris$Species), main = class, col = 1:3)) %>%
   invisible   # pipe to invisible to suppress output
```




# The same via `ggplot2`
```{r}
library(ggplot2)
library(reshape2)
thm <- theme_bw() +
  theme(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)

df <- melt(iris, id.vars = "Species", variable.name = "Feature",
           value.name = "cm") %>%
  mutate(Feature=gsub("\\.", " ", Feature))

ggplot(df, aes(y = cm, x = Species, fill = Species)) +
  geom_boxplot(color = "#1F3552", alpha=0.75, size=0.5) +
  scale_x_discrete(name="Species") +
  ggtitle("Overall Title") +
  theme(plot.title = element_text(hjust=0.5)) +
  facet_wrap(~Feature, ncol = 2) 
```




# Pairwise plots
Very useful for visually inspecting variables for patterns of interest. You will typically need to know your outcome of interest *a priori*, i.e. supervised analysis, or you will not know how to color the points.

```{r}
plot(iris[, -5], col=iris$Species)
```


```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2)
```



-----------------------------

