---
title: 'STAA 577: Laboratory Seven (II) </br> `tidyverse` version'
author: 'Adapted by Tavener & Field </br> From: James, Witten, Hastie and Tibshirani'
date: "`r format(Sys.Date(), '%e %B %Y')`"
output:
  html_notebook: 
    code_folding: show
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: no
ratio: '9:16'
tables: yes
fontsize: 12pt
---



```{r setup, include = FALSE}
options(warnPartialMatchArgs = FALSE)  # don't want these warnings
library(magrittr)     # pipes
library(tibble)       # tibbles
library(dplyr)        # data wrangling
library(purrr)        # iteration
library(ggplot2)      # tidy plotting
library(ISLR)         # Wage data set
library(splines)      # fitting splines
library(gam)          # General Additive Models
```



```{r data}
Wage %<>% as.tibble()     # convert to tibble
names(Wage)               # See variables available
Wage                      # look at tibble to see data
age      <- Wage$age      # make variable globally available downstream
wage     <- Wage$wage     # make variable globally available downstream
agelims  <- range(age)
age_grid <- seq(from = agelims[1], to = agelims[2])  # sequence from min to max
wage %>%                  # plot the distribution of the primary response variable of this lab
  data.frame() %>%
  ggplot2::ggplot(aes(x = wage)) +
  geom_histogram(binwidth = 5)
```

# Splines
## B-splines
Fit b-spline (by default a cubic spline) and plot spline +/- 2*standard
error bounds:

```{r splines}
b_spline_fit <- stats::lm(wage ~ splines::bs(age, knots = c(25, 40, 60)),
                          data = Wage)
bs_pred      <- predict(b_spline_fit,
                        newdata = data.frame(age = age_grid),
                        se = TRUE)

# df = (number of degrees of freedom) locates knots 
# based on uniform percentiles of the data
dim(bs(age, knots = c(25, 40, 60)))
dim(bs(age, df = 6))
attr(bs(age, df = 6), "knots")

# Fit a natural spline with df = 4
ns_spline_fit  <- stats::lm(wage ~ splines::ns(age, df = 4), data = Wage)
ns_pred        <- predict(ns_spline_fit,
                          newdata = data.frame(age = age_grid),
                          se = TRUE)
```


Calculate predicted values and standard errors (se), then collect
into a data frame for plotting with `ggplot`:
```{r plot_fitted_splines}
gg_df <- data.frame(      # gather relevant vectors into a data frame
  age_grid = age_grid,
  bs_fit   = bs_pred$fit,
  ns_fit   = ns_pred$fit
  ) %>% 
  dplyr::mutate(bs_upper = bs_fit + 2 * bs_pred$se.fit,  # calc 2se
                bs_lower = bs_fit - 2 * bs_pred$se.fit) %>%  # calc 2se
  tibble::as.tibble()

gg_df    # quickly view the fitted plotting data for lines

# pass full `x` and `y` data as df to ggplot
data.frame(age = age, wage = wage) %>%
  ggplot(aes(x = age, y = wage)) +
  geom_point(alpha = 0.25) +
  geom_line(data = gg_df, aes(x = age_grid, y = bs_fit), col = "blue") +
  geom_line(data = gg_df, aes(x = age_grid, y = bs_upper),
            col = "red", linetype = "dashed") +
  geom_line(data = gg_df, aes(x = age_grid, y = bs_lower),
            col = "red", linetype = "dashed") +
  geom_line(data = gg_df, aes(x = age_grid, y = ns_fit), col = "black") +
  NULL
```


------------------------


## Smoothing Splines
  
```{r smoothing_splines}
# Fit a smoothing spline
smooth_fit <- stats::smooth.spline(age, wage, df = 16)

# Select level of `smoothness` by cross-validation
smooth_fit_cv <- stats::smooth.spline(age, wage, cv = TRUE)
smooth_fit_cv$df     # ~7 degrees of freedom for the final spline

# create a new reorganized df of fitted values for plotting below
smooth_df <- data.frame(
  x         = smooth_fit$x,
  smooth    = smooth_fit$y,
  smooth_cv = smooth_fit_cv$y
  ) %>% 
  tidyr::gather(key = "fit", value = "predicted", -x) %>% 
  tibble::as.tibble()

smooth_df    # quickly view the plotting data

full_df <- data.frame(age = age, wage = wage)

full_df %>% 
  ggplot(aes(x = age, y = wage)) +
  geom_point(alpha = 0.25) +
  ggtitle("Smoothing Spline") +
  geom_line(data = smooth_df, aes(x = x, y = predicted, colour = fit)) +
  scale_colour_manual(values = c("red", "blue"), name = "") +
  theme(legend.position = "top") +
  NULL
```

--------------------------




# General Additive Models (GAMs)
## Fitting and plotting
Sum of two natural splines and a linear function:

```{r GAM_fit_plot, warning = FALSE, fig.height = 9/2.5, fig.width = 9}
# Fit a GAM using natural spline (`splines::ns`)
gam_ns <- stats::lm(wage ~ splines::ns(year, 4) + splines::ns(age, 5) + education,
                    data = Wage)

# Fit GAM using smoothing spline (`gam::s`)
# Sum of two splines and a linear function
gam_m3 <- gam::gam(wage ~ s(year, 4) + s(age, 5) + education,
                   data = Wage)

# There is a S3 method for the `plot` generic
# for objects of class `Gam`
# We'll make use of it here in lieu of developing our own `ggplot`
# Exercise: generate your own plot method in `ggplot`?
par(mfrow = c(1, 3))
plot(gam_m3, se = TRUE, col = "navy")

class(gam_ns)     # `gam_ns` is actually a `lm` object, but the `Gam` method recognizes it

# but we have to call the S3 method explicitly
plot.Gam(gam_ns, se = TRUE, col = "red")
```



## Compare, summarize, and predict
Fit two additional models and compare to the original `gam_m3` using ANOVA.
The models are:

  * **gam_m1**: excludes `year` as covariate
  * **gam_m2**: includes `year` as a *linear* covariate
  * **gam_m3**: includes `year` as a *non-linear* spline function (`df = 4`)

```{r GAMs_compare_summarize, warning = FALSE}
gam_m1 <- gam::gam(wage ~ s(age, 5) + education, data = Wage)
gam_m2 <- gam::gam(wage ~ year + s(age, 5) + education, data = Wage)
anova(gam_m1, gam_m2, gam_m3, test = "F")

# Based on the ANOVA, `gam_m2` is preferred
# Use `summary` generic to view `gam_m3`
# THERE IS SOMETHING WRONG HERE; CANNOT REPRODUCE THE
# VALUES IN THE TEXT BECAUSE THE SUMMARY METHOD HAS CHANGED
# ALSO, THEY SAY M2 IS PREFERRED, THEN PROCEED TO SUMMARIZE M3, WHY?
summary(gam_m3)

# Don't forget about `broom::tidy` 
# Easily and consistently summarize model output
broom::tidy(gam_m3)

# Use the built-in S3 `Gam` method to the `predict` generic
# to make predictions on the training set & calc MSE
# The `predict()` generic lives in stats pkg
# It is not necessary to select the model covariates;
# it will happen internally inside `predict`
# But it can be a good practice anyhow, to ensure you're not accidentally cheating
# Below use `dplyr::select` to keep only the predictors in the "new data"
pred_wage <- predict(gam_m2, newdata = dplyr::select(Wage, year, age, education))  
mse   <- mean((wage - pred_wage)^2)   # calculate MSE
mse
```
