---
title: "Simple rsample Basics"
author: "Stu Field"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output:
  html_notebook: default
ratio: '9:16'
tables: yes
fontsize: 12pt
---


## Contents
1)  Create a simple bootstrap
2)  Create LOO cross-validation
3)  Create a simple 10-fold cross-validation


## Setup
Load necessary libraries and recase `iris` dataset:
```{r, results = FALSE}
library(tibble)
library(magrittr)
library(purrr)
library(rsample)
iris %<>% as.tibble()
```

Look at the basic functionality of `rsample`:
``` r
library(help = rsample)
```


## Bootstrap sampling (out-of-bag; OOB):
Create some bootstrap sample sets from the `iris` data set
```{r}
set.seed(101)
boot5 <- rsample::bootstraps(iris, times = 5)
class(boot5)
boot5
```
#### Notice:

  * that `boot5` is a special class object (`bootstraps`, `rset`) but it's also a `tibble` object, so on it's base it's just a classy data frame. 
  * We are used to seeing data frames contain scalars or single elements in each of it's n x p cells, however doesn't *have* to be the case. The `splits` column contains special `rsplit` objects. If you look closely at them with `str`, you'll see that they themselves are just lists, which themselves contain data frames and vectors of the indices for the samples that are included in the bootstrap.
  * It's like a russian nesting doll of data!
  * The advantage once this structure exists is that all one must do is iterate over the rows of the tibble (`boot5`) to proceed with the model fitting process. All the necessary information is contained in each `splits` column of the tibble.
  * Conveniently, the `tidyverse` has a package called `purrr` which is full of specialized functions designed to make **iteration** easy.



Proportion *virginica* species in original data:
```{r}
mean(iris$Species == "virginica")   # 1/3
```

Use the `purrr` package suite of iteration flavors of `map` to look at the proportion *virginica* by bootstrap "population":
```{r}
purrr::map_dbl(boot5$splits,
  function(x) {
  species <- as.data.frame(x)$Species
  mean(species == "virginica")  # proportion shorthand
})
```


## Leave-one-out cross-validation (LOOCV)
Special case of k-fold cross-validation:

```{r}
set.seed(101)
looCV <- rsample::loo_cv(iris)
class(looCV)
looCV
```

Find the missing (left-out) sample by iteration:
```{r}
purrr::map_dbl(looCV$splits,
  function(x) {
  setdiff(1:nrow(x$data), x$in_id)
})
```

## Cross-Validation (CV)
* k-fold cross-validation
    * aka v-fold cross-validation
    * aka x-fold cross-validation

Generate 10-fold cross-validation data sets and store in object `cv10`
```{r}
set.seed(101)
cv10 <- rsample::vfold_cv(iris, v = 10)
dim(iris)   # 150 samples; 4 Features; 1 Class name "Species"
```

There *should* be 135 vs 15 samples in each of 10 folds, let's check:
```{r}
purrr::map_df(cv10$splits,
  function(x) {
  n_in  <- x$in_id
  n_out <- setdiff(1:nrow(x$data), n_in)
  fold  <- x$id %>% as.character()
  data.frame(fold_id = fold,
             samples_in = length(n_in),
             samples_out = length(n_out))
})
```

Created on `r Sys.Date()` by the [reprex package](http://reprex.tidyverse.org) (v`r utils::packageVersion("reprex")`).
